The effect of habitat choice on evolutionary rescue in subdivided populations

### 2020-06-08 ms/main.tex
revision en cours

\usepackage[utf8x]{inputenc}	
\usepackage{fullpage}		
\usepackage[T1]{fontenc}				
\usepackage[english]{babel}			
%\usepackage{ulem}
\usepackage{xcolor}						
\usepackage{amsmath,amstext,amssymb,amsfonts,amsthm,mathrsfs,chemarrow,stackrel,dsfont}   % Math packages
\usepackage{graphicx} 					
\usepackage{subcaption}					
%\usepackage{cite}	
\usepackage[authoryear,sectionbib,sort]{natbib} % Bio citation format
\linespread{1.7}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

@@

\usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref} 
\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
\usepackage{FiraSans}
\usepackage{soul}

% Packages to comment out before submitting!
%\usepackage[textsize = scriptsize, backgroundcolor = white, linecolor = black]{todonotes}
%\usepackage{showlabels}

@@

\noindent The authors wish to be identified to the reviewers.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and empirical evidence %%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Facing current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, the populations of many species are declining and might eventually go extinct \citep{bellard_2012, diniz_2019}. Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}. 
The process of genetic adaptation that saves populations from extinction is termed evolutionary rescue. This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, typically }resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}. Evolutionary rescue \linelabel{R1-4}\chg{has been observed directly in laboratory experiments \citep[e.g.][]{bell_2009,agashe_2011,lachapelle_2012,lindsey_2013,stelkens_2014}. In wild populations demographic and genotypic data is rarely monitored at the same time impeding direct observation of evolutionary rescue. Still, it has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012,digiallonardo_2015,gignoux_2018}.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and horizontal gene transfer \citep{tazzyman_2014}, reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\chg{\citep{yamamichi_2015}}, bottlenecks \citep{martin_2013}, different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of old- and new-habitat patches until eventually the whole environment has deteriorated. If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}. Experiments that study the effect of dispersal on evolutionary rescue are rare but indicate that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation \citep{bell_2011}.

\linelabel{R1-1}\chg{Evolutionary rescue in a fragmented environment can be interpreted as the intermediate situation between abrupt change and gradual change: if dispersal is very low, patches are approximately isolate and thus each patch undergoes an abrupt change independent of the other patches; if the dispersal rate and the number of patches are very high, the whole environment is gradually changing since spatial structure and the contribution of a single patch are negligible.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Adaptation and dispersal %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Evolutionary rescue in a situation where one patch after \linelabel{R2-2}\chg{another} deteriorates is tightly linked to the study of adaptation to a heterogeneous environment with source-sink dynamics. These describe a spatially heterogeneous environment that is constant in time where a population in unfavorable habitats can be maintained by constant immigration of the wild type. Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g. \citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000,ronce_2001,gallet_2018}} for both effects).
%Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)
\florence{suggestion: focus on discrete traits - thousands of possible papers to cite; avoid RK2001}

In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be \chg{unbiased}, i.e. dispersing individuals are distributed uniformly among patches. Only few investigations in the context of local adaptation in source-sink systems have taken into account non-\chg{uniform} dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}. This analytical focus on \chg{unbiased} dispersal is in stark contrast to actually observed dispersal schemes in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}. Here, we explore and compare the effects of several biologically motivated dispersal schemes on adaptation and evolutionary rescue.

@@

Another well-established dispersal scheme is habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to. This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.

\linelabel{R1-8}\chg{Although density-dependence or habitat choice can also affect the emigration behavior, i.e. the likelihood that an individual leaves its current location, or even the vagrant stage when the individual is in the process of dispersal \citep{bowler_2005,ronce_2007}, we ignore this here. Instead, we concentrate on \chg{non-uniform} dispersal effects of the immigration process.} \\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What do we do? %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In the following, we explicitly account for these \chg{biased} dispersal schemes in the context of genetic adaptation to a deteriorating and spatially structured environment. We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}\chg{types, wild type and mutant, have a positive growth rate}, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct. In this framework, we study four biologically motivated dispersal patterns and compare them to the \chg{unbiased} dispersal scheme. We start by investigating their consequences for the establishment dynamics of an adapted mutant in a time-constant heterogeneous environment. 
From there, we derive an approximation for adaptation in a source-sink setting.
Using these results, we study the probability of evolutionary rescue, i.e. of adaptation in the scenario where patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat. We find that dispersal bias has a direct effect on the local growth rates which in turn affect the probability of establishment and evolutionary rescue.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% MODEL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. The transition from the old to the new habitat is assumed to be irreversible. Patches are populated by two types of asexually reproducing, haploid individuals, either wild type or mutant. Generations are measured in discrete time steps and are non-overlapping. \linelabel{R1-9-2}\linelabel{R2-6}\chg{Wild-type and mutant individuals both have a positive growth rate} in old-habitat patches; the mutant has a lower growth rate there. In new-habitat patches, the wild type population declines locally while the mutant has a positive growth rate -- we will therefore call it a `rescue mutant'. \chg{Both habitat types, old and new, have a (potentially different) carrying capacity. At the end of a life cycle, the local population size is reduced to the carrying capacity if necessary.}

We consider a time-varying environment, such that one after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates (from old to new state). Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event until it eventually hits~$0$, when all patches have undergone the environmental change. This environmental setting corresponds to the one analyzed by \citet{uecker_2014} and more recently in the special case of just two patches by \citet{tomasini_2019}.

The individuals are assumed to go through the following life-cycle: (i) Dispersal: individuals may move between patches; (ii) Reproduction: individuals reproduce within patches. \linelabel{R2-7}\chg{The number of offspring per parent is modeled by a Poisson number}; (iii) Mutation: wild-type individuals mutate to the rescue mutant type with probability $\theta$ -- back mutations from the mutant to the wild type are neglected; (iv) Regulation: the population size is, \chg{if necessary, down-regulated to the carrying capacity $K_k$ where $k$ refers to the habitat type (old or new - the two habitat types can have different carrying capacities)}. If the population size is below the carrying capacity, the regulation step is ignored.

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Old habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Old-habitat dynamics}
\linelabel{R1-10}We assume that in old-habitat patches, \linelabel{R1-9-3}\chg{both types have a positive growth rate such that the population in these patches is typically at carrying capacity $K_{\text{old}}$.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega_w$ and $\omega_m$, respectively. 
We impose an adaptation trade-off for the rescue mutant, i.e. the mutant is less fit than the wild type in old-habitat patches ($\omega_m<\omega_w$).
%
\linelabel{R2-9}\chg{We assume that mutation from the wild type to the mutant is tied to reproduction. The probability that an offspring of a wild-type individual is a mutant is given by $\theta$. We neglect back-mutation from mutant to wild type.} 
\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e. we assume that the wild type and the mutant are competitively equivalent. The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by   
\begin{equation}\label{eq:sold}
     K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}^{\text{old}}_w + \omega_m}\approx K_{\text{old}}\, \frac{\omega_m}{\omega_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
\end{equation}
where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step. Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$. \linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e. $\omega_m<\omega_w$. Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small. This strongly depends on the dispersal scheme and rate. \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}

@@

\begin{itemize}
	\item \textbf{Old-Old:} Individuals prefer to immigrate into the habitat where they have the largest number of offspring before regulation. \linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega_w>1-r$ and $\omega_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e. a bias towards old-habitat patches. In practice, individuals are thought to use biotic or abiotic cues to preferentially immigrate into habitats where their fecundity is highest when compared to other habitats. This type of dispersal, i.e. matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.

@@

@@

\chg{In line with this reasoning we find} that the width of region (ii) strongly depends on the fecundity of the mutant in the old habitat, $\omega_m$. If $\omega_m$ is high, the local growth rate of the mutant $a_{\text{old}}$ starts to increase at a lower dispersal rate 
%(see Fig.~S1). 
\chg{and }region (ii) disappears, as visible in Fig.~\ref{fig:vary_m_est}(c). In contrast, for low mutant fecundity values $\omega_m$, region (iii) might vanish (see Fig. S2 in SI). There, the mutant is strongly disadvantageous in old-habitat patches when compared to the wild type, and the wild-type population will always outcompete the mutant due to the much higher offspring numbers.

>>>>

\usepackage[utf8x]{inputenc}	
\usepackage{fullpage}		
\usepackage[english]{babel}			
%\usepackage{ulem}
\usepackage{xcolor}						
\usepackage{amsmath,amstext,amssymb,amsfonts,amsthm,mathrsfs,chemarrow,stackrel,dsfont}   % Math packages
\usepackage{graphicx} 					
\usepackage{subcaption}					
%\usepackage{cite}	
\usepackage[authoryear,sectionbib,sort]{natbib} % Bio citation format
\linespread{1.7}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

\usepackage{enumitem}

@@

\usepackage[pdftex, breaklinks, colorlinks, citecolor=black, urlcolor=black, linkcolor=black]{hyperref} 
\usepackage{fourier} % FD I try to avoid the default LaTeX font for non math journals
\usepackage{FiraSans}
\usepackage[T1]{fontenc}

\usepackage{soul}

% Packages to comment out before submitting!
\usepackage[textsize = scriptsize, backgroundcolor = white, linecolor = black]{todonotes}
%\usepackage{showlabels}

@@

\noindent The authors wish to be identified to the reviewers (the manuscript is already on bioRxiv so the identities of all authors can be easily found).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and empirical evidence %%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Current anthropogenic environmental changes such as deforestation, soil and water contamination or rising temperatures, contribute to the decline of the populations of many species, that might eventually go extinct \citep{bellard_2012, diniz_2019}. Pests and pathogens experience similarly strong selective pressures as a result of increased consumption of antibiotics and use of pesticides \citep{ramsayer_2013, kreiner_2018}. 
The process of genetic\footnote{\florence{rescue by plasticity as well}} adaptation that saves populations from extinction is termed evolutionary rescue. This \linelabel{R1-3}\chg{process is characterized by an initial population decline (that would result in population extinction) followed by recovery due to the establishment of adapted genotypes, } \florence{classically} resulting in a U-shaped demographic trajectory over time~\citep{gomulkiewicz_1995}.
In recent years, empirical examples of evolutionary rescue have accumulated \citep[as reviewed by ][]{alexander_2014,carlson_2014,bell_2017}. Laboratory experiments have provided direct evidence of evolutionary rescue \linelabel{R1-4} \citep[e.g.][]{bell_2009, agashe_2011, lachapelle_2012, lindsey_2013, stelkens_2014}. In the wild, however, demographic and genotypic data are rarely monitored together at the same time, which impedes direct observation of evolutionary rescue. Still, evolutionary rescue has been suggested as a mechanism that has saved a few wild populations from extinction \citep[e.g.][]{vanderwal_2012, digiallonardo_2015, gignoux_2018}.

%% Genetic setting

\florence{[Need a paragraph on the genetic makeup. We are not explaining this at all at the moment and this is lacking] Rescue can occur through plasticity (refs Chevin? \footnote{https://royalsocietypublishing.org/doi/full/10.1098/rspb.2016.1690 , https://royalsocietypublishing.org/doi/abs/10.1098/rstb.2012.0089 and/or https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2009.00875.x}) or genetic adaptation (refs). The traits involves can be continuous (ref) or discrete (ref). In this work, we consider genetic adaptation mediated by discrete traits. Specifically, we consider that a single mutation can prevent extinction, and...}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Evol rescue and dispersal %%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In mathematical models, evolutionary rescue is often studied in a spatially homogeneous situation where the whole population experiences a sudden decrease in habitat quality. In this setting, a large number of theoretical results have been established, for example on the effects of recombination \citep{uecker_2015} and horizontal gene transfer \citep{tazzyman_2014}, reproduction mechanisms \citep{glemin_2013,uecker_2017}, intra- and interspecific competition \citep{osmond_2013}, predation pressure \linelabel{R1-5}\chg{\citep{yamamichi_2015}}, bottlenecks \citep{martin_2013}, different genetic pathways \citep{osmond_2019}, and the context-dependent fitness effects of mutations \citep{anciaux_2018}. \chg{In contrast to these abrupt change scenarios, evolutionary rescue can also be studied in a gradually changing environment \citep[e.g.][]{osmond_2017}}\footnote{\pete{Does anybody know other references?} \hildegard{What about Lynch et al. 1991. Adaptive and  demographic responses of plankton populations to environmental change. Limnology and Oceanography 36:1301--1312., B{\"u}rger and Lynch 1995. Evolution and extinction in a changing environment: a quantitative genetic analysis. Evolution 49:151--163 and Lande and Shannon 1996. The role of  genetic  variation  in adaptation and population persistence in a changing environment. Evolution 50:434--437? But please check; it's a long time that I read those...}}. 
Such gradual changes can in particular occur in fragmented environments.

In fragmented environments, habitat deterioration is not necessarily synchronized across patches: there can be a transient spatially heterogeneous environment consisting of a mosaic of old and of degraded habitat patches, until eventually the whole environment has deteriorated. If individuals that populate different patches are able to move between those, the effect of dispersal on evolutionary rescue needs to be taken into account \citep{uecker_2014, tomasini_2019}. \linelabel{R1-1}\chg{The intensity of dispersal among patches tunes how abruptly environmental change is experienced. With very low dispersal, patches are essentially isolated from each other, and each patch undergoes an abrupt change independently of the other patches. With higher dispersal, asynchronous deterioration among patches is experienced as a more gradual change overall.}
Experiments that study the effect of dispersal on evolutionary rescue are rare, but, for instance, \citet{bell_2011} found that \linelabel{R1-6} dispersal can increase the likelihood of successful genetic adaptation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Adaptation and dispersal %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transient mosaic of degraded and non-degraded patches that results from asynchronous degradation in a fragmented habitat is similar to the setting of models of source-sink dynamics. These models represent a spatially heterogeneous environment, constant in time, in which wild-type populations in unfavorable habitats can only be maintained thanks to recurrent immigration from favorable habitats. 
Experimental and theoretical studies have found that \linelabel{R2-3}\chg{increasing} dispersal can have a positive or a negative effect on genetic adaptation in a heterogeneous environment (see e.g., for studies on discrete traits, \florence{[refocus the citations with papers on discrete traits; otherwise there are too many possible citations -- it's fine to do so given that we now explicitly say the paper is focused on discrete traits]}\citet{holt_1997,gomulkiewicz_1999} for positive; \citet{storfer_1998,garcia_1997,kirkpatrick_1997,fedorka_2012} for negative; and \linelabel{R1-7}\chg{\citet{kawecki_2000, gallet_2018}} for both effects).
%Reason for different results: (i) demography or not (Kawecki); (ii) selection strength (Gallet), (iii) selection strength (Ronce)

%%% DISPERSAL, biased or not
In theoretical studies of local adaptation and evolutionary rescue, dispersal is typically assumed to be \chg{unbiased}, i.e. dispersing individuals are distributed uniformly among patches. Only few investigations in the context of local adaptation in source-sink systems have taken into account non-\chg{uniform} dispersal patterns \citep[e.g.][]{kawecki_1995,holt_1996,kawecki_2002,amarasekare_2004}. This analytical focus on \chg{unbiased} dispersal is in stark contrast to dispersal schemes observed in nature \citep[][]{edelaar_2008,clobert_2009,edelaar_2012}.

@@

Another well-established dispersal scheme is a type of habitat choice, whereby individuals tend to immigrate into habitats they are best adapted to. This mechanism has for example been reported in lizards \citep{bestion_2015}, birds \citep{dreiss_2011,benkman_2017}, fishes \citep{bolnick_2009}, worms \citep{mathieu_2010}, and ciliates \citep{jacob_2017,jacob_2018}.

\linelabel{R1-8} Dispersal biases can affect the different steps of dispersal (the probability to emigrate, the vagrant stage, and immigration \citep{bowler_2005,ronce_2007}). In this work, we focus on effects on the immigration step.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% What do we do? %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We model an environment that consists of various patches with one of two possible habitats: the `old' habitat, in which both \linelabel{R1-9-1}\chg{types, wild type and mutant, have a positive growth rate}, and the `new' habitat, where in the absence of immigration the wild-type population will eventually go extinct. We study four biologically motivated dispersal schemes, which correspond to the four combinations of biases towards old vs.\ new patches for wild type and mutants, and we compare these dispersal schemes to \chg{unbiased} dispersal. Our analysis is carried out step-wise. We first consider a temporally constant but spatially heterogeneous environment with two (`old' and `new') patch types. In this setting, we first study the probability of establishment of a single mutant, assuming there are no further mutations between types. We then relax the assumption of no further mutations, and compute a probability of adaptation, i.e. of establishment of the mutant lineage\footnote{\florence{not sure about this defn, please check}}. Finally, we let habitat degradation proceed, assuming that patches, one after \chg{another}, deteriorate over time until all locations contain the new habitat. Using the previous results, we approximate a probability of evolutionary rescue, i.e.\ that a mutation appears, establishes, thereby allowing the population to persist in spite of environmental degradation. We find that dispersal biases affect the probabilities of establishment and of evolutionary rescue, because it directly affects the local growth rates.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%% MODEL %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%
% General description %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a spatially structured environment consisting of $M$ patches all connected to each other. The habitat of a patch is either in the \textit{old} or in the \textit{new} state, corresponding to habitat quality before and after environmental deterioration, respectively. One after \chg{another} every $\tau$ generations, the habitat of a patch deteriorates, from old to new state, the transition being irreversible. Initially ($t<0$), all patches are of the old-habitat type. At time $t=0$, the first patch deteriorates. After $(M-1)\tau$ generations, all patches are of the new-habitat type. We denote the time-dependent frequency of old-habitat patches by $f_{\text{old}}$. It equals $1$ before the first environmental change takes place ($t<0$), and decreases by $1/M$ after each environmental deterioration event, until it eventually hits~$0$, when all patches have undergone the environmental change. This setting corresponds to the one analyzed by \citet{uecker_2014}, and more recently by \citet{tomasini_2019} in the special case of just two patches\footnote{\florence{what about carrying capacities in these papers?}}. The maximum numbers of individuals that can live in a patch of a given habitat type, i.e.\ the carrying capacities, are denoted $K_{\text{old}}$ and $K_{\text{new}}$ for old- and new-habitat patches respectively; $K_{\text{old}}$ and $K_{\text{new}}$ may differ.

The population living in this environment consists of asexually reproducing, haploid individuals; generations are discrete and non-overlapping. There are two possible types of individuals, wild types and mutants. The individuals go through the following life-cycle: \begin{enumerate}[label = (\roman*)]
	\item Dispersal: individuals may move between patches. Further details about this step are given below.
	\item Reproduction: individuals reproduce within patches. \linelabel{R2-7}The number of offspring produced by an individual of type $i$ in habitat $k$ (before density regulation, if any), i.e.\ its fecundity, is drawn in a Poisson distribution of expectation $\omega_i^k$. Having fewer than $1$ offspring in expectation means that the local subpopulation will get extinct in the absence of immigration, because the deaths of the parents at each generation are not compensated by enough births on average. On the contrary, the local population is viable if the expected fecundity is greater than $1$. We assume that wild-type and mutant populations are viable in old-habitat patches, and that the mutant's expected fecundity there is lower than the wild type's: $1 \leq \omega^{\text{old}}_m < \omega^{\text{old}}_w$ \footnote{\florence{In the table, it is $0 \leq $, yet the populations are supposed to be viable}}. In new-habitat patches, a wild-type population will eventually go extinct, while a mutant one would persist, hence the term ``rescue mutant'': $\omega^{\text{new}}_w < 1 < \omega^{\text{new}}_m$. All parents die.
	\item Mutation: wild-type offspring mutate to the rescue mutant type with probability $\theta$ (back mutations from the mutant to the wild type are neglected).
	\item Regulation: if the number of offspring produced locally exceeds the local carrying capacity $K_k$ (where $k$ refers to the habitat type, old or new), the population size is down-regulated to the $K_k$: individuals are randomly removed until the local population size is equal to $K_k$. One mutant offspring has the same chance of being removed as one wild-type offspring: \linelabel{AE-3} we assume that wild-type and mutant individuals are competitively equivalent. If the number of offspring is below the carrying capacity, the regulation step is ignored. 
	At the end of this step, all offspring become adults, and a new cycle then begins.
\end{enumerate}

@@

\begin{itemize}
	\item \textbf{Old-Old:} Individuals prefer to immigrate into the habitat where they have the largest number of offspring before regulation. \linelabel{R1-22}\linelabel{R2-16}\chg{Since the average number of offspring per parent is higher in the old habitat for both types ($\omega^\text{old}_w>1-r$ and $\omega^\text{old}_m>1+a_{\text{new}}$), this translates} to both $\pi_w$ and $\pi_m$ being larger than~$0$, i.e. a bias towards old-habitat patches. In practice, individuals are thought to use biotic or abiotic cues to preferentially immigrate into habitats where their fecundity is highest when compared to other habitats. This type of dispersal, i.e. matching habitat choice, has for example been observed with common lizards \textit{Zootoca vivipara} \citep{bestion_2015}, three-spine sticklebacks \textit{Gasterosteus aculeatus} \citep{bolnick_2009}, and barn owls \textit{Tyto alba} \citep{dreiss_2011}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%
% Old habitat dynamics %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Old-habitat dynamics}
\florence{It is not clear whether population size is at Kold or not}
\linelabel{R1-10}\chg{We assume that the populations in old-habitat patches \linelabel{R1-9-3}are at carrying capacity $K_{\text{old}}$, since both types have a positive growth rate in these patches.} We denote the fecundities of the \textbf{w}ild-type and the \textbf{m}utant by $\omega^\text{old}_w$ and $\omega^\text{old}_m$, respectively. We assume that adaptation to the new habitat comes at a cost, such that the mutant is less fit than the wild type in old-habitat patches ($\omega^\text{old}_m<\omega^\text{old}_w$).
%
\linelabel{R2-9}Mutations occur upon reproduction\footnote{\florence{please check my phrasing}}. The probability that the offspring of a wild-type individual is mutant is given by $\theta$. We neglect back-mutation from mutant to wild type.
\linelabel{R2-8}\chg{After reproduction, if the population size exceeds the carrying capacity $K_{\text{old}}$ we randomly remove individuals from the population until the carrying capacity is reached\linelabel{AE-3}, i.e. we assume that the wild type and the mutant are competitively equivalent. The mean number of offspring of\footnote{\francois{produced by}} one mutant after reproduction and population regulation is then given by   
\begin{equation}\label{eq:sold}
     K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w + \omega^\text{old}_m}\approx K_{\text{old}}\, \frac{\omega^\text{old}_m}{\omega^\text{old}_w \widetilde{N}^{\text{old}}_w} =1 + a_{\text{old}} \, ,
\end{equation}
where the random variable $\widetilde{N}_{i}^{k}$ denotes the population sizes of type $i$ (mutant or wild-type) in patches with habitat $k$ (old or new) after the dispersal step. Additionally, we have defined the local growth rate of a single mutant, denoted by $1+a_{\text{old}}$. \linelabel{R1-11}The approximation in eq.~\eqref{eq:sold} holds if the number of mutants is much lower than the number of wild types in old-habitat patches -- a reasonable assumption since the establishment of the mutant is decided when it is still rare.}
%Then, while still rare, mutant individuals will reproduce independently of each other and their dynamics will be described by the local growth rate given in eq.~\eqref{eq:sold}. 
\chg{Compared to the wild type we assume a reduced fecundity of the mutant in the old habitat, i.e. $\omega^\text{old}_m<\omega^\text{old}_w$. Still, the local growth rate can be positive, that is, larger than one ($a_{\text{old}}>0$).} This happens when the \linelabel{R2-12}\chg{wild-type population in a single old-habitat patch after dispersal,} \linelabel{R1-13}$\widetilde{N}^{\text{old}}_w$, becomes very small. This strongly depends on the dispersal scheme and rate. \linelabel{R1-14}\chg{The population dynamics of the wild type are analyzed explicitly in Section~1 of the Supplementary Information (SI).}

@@

\chg{In line with this reasoning we find} that the width of region (ii) strongly depends on the fecundity of the mutant in the old habitat, $\omega^\text{old}_m$. If $\omega^\text{old}_m$ is high, the local growth rate of the mutant $a_{\text{old}}$ starts to increase at a lower dispersal rate 
%(see Fig.~S1). 
\chg{and }region (ii) disappears, as visible in Fig.~\ref{fig:vary_m_est}(c). In contrast, for low mutant fecundity values $\omega^\text{old}_m$, region (iii) might vanish (see Fig. S2 in SI). There, the mutant is strongly disadvantageous in old-habitat patches when compared to the wild type, and the wild-type population will always outcompete the mutant due to the much higher offspring numbers.
